---
title: 'Reproducible Research: Peer Assessment 1'
author: "Yuva"
date: "July 7, 2016"
output: html_document
---
Reproducible Research:Peer Assessment 1
==========================================
Prepare the R environment
=========================
Throughout this report when writing code chunks in the R markdown document, always use echo = TRUE so that someone else will be able to read the code.

First, we set echo equal to TRUE and results equal to 'hold' as global options for this document.
```{r}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```
load required libraries
```{r}
library(data.table)
library(ggplot2)
```
loading the unzipped data file into R
===================================
```{r}
activity <- read.csv('activity.csv', header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric"))
```
preprocessing the loaded data
===============================
convert the date field to Date class and interval field to Factor class.
```{r}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
activity$interval <- as.factor(activity$interval)
```
lets check the structure of the data using str()
===============================
```{r}
str(activity)
```
Mean total number of steps per day?
==================================
calculation total steps per day
```{r}
steps_per_day <-aggregate(steps ~ date, activity, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```
1. Now we make a histogram of the total number of steps taken per day, plotted with appropriate bin interval.
```{r}
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Frequency") + theme_bw()
```

Mean and median of total number of steps per day
==================
```{r}
mean_steps <- mean(steps_per_day$steps, na.rm = TRUE)
median_steps <- median(steps_per_day$steps, na.rm=TRUE)
mean_steps
median_steps
```
mean steps per day = 10766.19
median steps per day = 10765

What is the average daily activity pattern?
==============================================
We calculate the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called steps_per_interval.
```{r}
steps_per_interval <- aggregate(activity$steps, 
                                by = list(interval = activity$interval),
                                FUN=mean, na.rm=TRUE)
## lets convert it to interger which will make the plot easy

steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```
We make a plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals
```{r}
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="blue", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()
```

we find the 5-minute interval with the containing the maximum number of steps
```{r}
steps_per_interval[which.max(  
        steps_per_interval$steps),]
```
The 835th interval has the maximum number of steps of 206

Imputting missing values
=========================
total number of missing values in the dataset

```{r}
sum(is.na(activity$steps))
```
Total number of missing values = 2304

Filling out missing value
====================
lets fill the missing values with the mean 
create a function na_fill(data, pervalue) which the data arguement is the activity data frame and pervalue arguement is the steps_per_interval data frame
```{r}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

activity_fill <- data.frame(  
        steps = na_fill(activity, steps_per_interval),  
        date = activity$date,  
        interval = activity$interval)
str(activity_fill)
```
lets check whether there is any missing value in the activity_fill data frame
```{r}
sum(is.na(activity_fill$steps))
```
Output is zero means there is no missing value in the dataframe

Histogram for the total number of steps taken per day
===========================
```{r}
fill_steps_per_day <- aggregate(steps ~ date, activity_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Frequency") + theme_bw() 
```

Calculate and report the mean and median total number of steps taken per day
==============================================================================

```{r}
mean(fill_steps_per_day$steps)
median(fill_steps_per_day$steps)
```

The mean and median of the data set is surprisingly equal to 10766.19. There is
not much of a difference between the original data and the data set filled with the mean for the missing value

Before filling the data
====================
Mean = 10766.19,

Median =10765

After filling the data
=====================
Mean = 10766.19,

Median = 10766.19

What is the impact of imputing missing data on the estimates of the total daily number of steps?
===========================================================

Both mean and median are almost identical. However, the data after filling the missing value shows almost normal distrbutions with a major peak around 12000 steps (higher than the one with the missing value) and and number of fequence for each steps is almost double after the missing values have been filled

Are there differences in activity patterns between weekdays and weekends?
=========================================================================

We do this comparison with the table with filled-in missing values.
1. Augment the table with a column that indicates the day of the week
2. Subset the table into two parts - weekends (Saturday and Sunday) and weekdays (Monday through Friday).
3. Tabulate the average steps per interval for each data set.
4. Plot the two data sets side by side for comparison.

```{r}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(activity_fill)
```
panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekend
```{r}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="blue") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```
 Week day has one major peak around 1000 interval but weekend has more peaks scattered thorough out the interval.It seems like people are active through out the Weekend and not in a specific time as in weekdays 










